### 2020-02-18 Writing/disparity_review.tex
Reformated until p.13

Statistical nalyses of morphological disparity have been used to
characterise and investigate the evolution of variation in the anatomy,
function, and ecology of organisms since the 1980s. While a diversity of
methods have been employed, it is unclear whether they provide
equivalent insights. Here we review the most commonly used approaches
for characterising and analysing morphological disparity, all of which
have associated limitations that, if ignored, can lead to
misinterpretation. We provide best practice guidelines for disparity
analyses, while noting that there can be no ``one-size-fits-all''
approach. The available tools should always be used in the context of a
specific biological question that will determine data and method
selection at every stage of the analysis.

@@

Clades of organisms are characterised by variation in both numbers of
species and range of phenotypes through time. At the extremes, clades
may be exceptionally species-rich and phenotypically diverse (e.g.
cichlids and molluscs), species-rich but phenotypically conservative
(e.g. bacteria and nematodes), species-poor but phenotypically diverse
(e.g. Afrotheria), or depauperate in both species and phenotypic
diversity (e.g. lungfish). These phenomena suggest that taxonomic and
phenotypic diversity are not inextricably linked, raising important
questions about how phenotypic diversity arises, such as: How does
diversity evolve? Are some morphologies more common than others? Can
anatomy evolve in all ``directions'' or are some anatomies impossible to
achieve? (Mike Foote 1997, 1995) What role does ecology play in
structuring morphological diversity?. Analyses of species diversity have
a venerable history, but those of phenotypic diversity (hereafter
\emph{disparity}) are a comparatively more recent phenomenon. Originally
defined as ``multidimensional morphological dissimilarity at a
macroevolutionary scale'' (Runnegar 1987, Gould
\href{https://paperpile.com/c/sTGYvp/Uns3}{2000}), this concept of
disparity emerged from attempts by palaeobiologists to characterise the
evolutionary origin of animal bodyplans and from attempts by comparative
developmental biologists to provide causal explanations for their
emergence. However, analyses of ``disparity'' have since expanded into
comparative biology as a means of capturing the effect of intrinsic and
extrinsic causal agents in morphological evolution. Typically, methods
to capture disparity are based on multidimensional spaces where each
dimension represents an aspect of morphological variation (a trait) and
biological observations (taxa) can be placed in this space based on
their trait values. Such multidimensional spaces (or morphospaces) can
then be used to tackle a diverse array of questions that can be grouped
into four main (non-mutually exclusive) classes (Fig. 1):

@@

(3) \emph{Disparity and taxonomic diversity.} Morphological disparity
provides another perspective on biodiversity; high morphological
disparity represents a high diversity of morphologies (i.e. shapes or
body plans) and is, presumably, associated with high levels of
ecological and functional diversity (though see below). This makes
disparity an informative complement to diversity measures based on
species richness alone. Indeed, most studies that have investigated
disparity and taxonomic diversity support an effective decoupling
\href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(e.g., Fortey,
Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and
Neige 2007)}. This approach has been used to investigate whether some
groups are more successful than others in their exploration of new
evolutionary strategies
\href{https://paperpile.com/c/sTGYvp/dJHu}{(Losos 2011)}.

@@

(5) Disparity and evo-devo. Morphological disparity blalbalbal

Fundamental insights into evolutionary biology have been elicited from
these four types of analysis. One of the most important insights is the
discovery that morphological disparity is often greatest early in the
evolutionary history of clades
\href{https://paperpile.com/c/sTGYvp/xxh5+yqPw}{(Hughes, Gerber, and
Wills 2013; Foote 1997;}
\href{https://paperpile.com/c/sTGYvp/EPJ2}{Erwin 2007)}, indicating that
the capacity for evolutionary innovation wanes with clade age, which
some have argued reflects the evolutionary assembly of gene regulatory
networks that constrain later fundamental change
\href{https://paperpile.com/c/sTGYvp/xxh5+Z6l6}{(Hughes, Gerber, and
Wills 2013; Erwin 2011)}. However, this example highlights one of the
most challenging problems confronting researchers who are attempting,
increasingly, to obtain general insights from multiple independent
studies: can the insights gained from studies using the diversity of
methods, approaches and data types employed be considered equivalent?

In attempting to answer this question, we review current methods and
highlight their limitations, as part of a more general attempt to
propose best practice guidelines for studies of disparity . We first
discuss the appropriate data required for characterising disparity, then
review various challenging aspects of these approaches. Throughout, it
is important to remember these tools should always be used in the
context of a specific scientific question, as this will drive data and
methodological choices at every stage of the process.

@@

@@

Disparity analyses are based on traits, but traits can be characterised
in a number of ways: (1) discrete ``cladistic'' characters, e.g. coding
the absence or presence of features or a discrete characteristic of a
trait \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)};
(2) continuous measurements of features (e.g., lengths)
\href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al.
2011)}; or (3) more mathematical descriptors from geometric morphometric
landmark data (e.g. Procrustes coordinates)
\href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, and
Fourier coefficients \href{https://paperpile.com/c/sTGYvp/2Neu}{(Foote
1989;} \href{https://paperpile.com/c/sTGYvp/ZEDR}{Spriggs et al. 2018)})
(Fig. 2). None of these approaches are superior, but they may be more or
less well-suited to characterising traits under comparison and to the
question being asked of those traits (Hetherington et al 2015; Hopkins
2017).

@@

The points above assume that researchers are collecting their own data
for disparity analyses, but often this is not the case. Discrete
character data are commonly recycled from phylogenetic studies
(e.g.Foote 1989\href{https://paperpile.com/c/sTGYvp/2Neu+0y4V}{; Deline
et al. 2018}). This is an efficient approach to character sampling, but
it may artifactually increase disparity between phylogenetically
distinct groups because phylogenetic characters are often collected to
discriminate among groups
\href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}. This needs to
be considered when interpreting results, especially as synapomorphies
will naturally lead to an apparent shift or increase in disparity when
new clades appear. Furthermore, many datasets are limited to subsets of
anatomy that are at least implicit samples of overall anatomy, but
explicit tests of this assumption have shown that different aspects of
morphology can exhibit different patterns of disparity
\href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}.
This effect of anatomical part on disparity patterns can be especially
challenging when working on datasets where the available data has
non-random missing anatomical parts, such as the absence of soft tissue
in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline
et al. 2018)}.

@@

Once suitable trait data have been collected, the design of the
disparity analysis itself needs to be considered. Study design
encompasses several key aspects including (a) the difficulty of dealing
with multidimensional data; (b) the variety of ordination (dimension
reduction) techniques available and their limitations; (c) the metrics
used to summarise the relative disparity of groups; (d) the methods used
for hypothesis testing within the disparity analysis framework; and (e)
ancestral state estimations in disparity-through-time analyses. We
consider these aspects in turn below.

>>>>

Analyses of morphological disparity have been used to characterise and investigate the evolution of variation in the anatomy, function, and ecology of organisms since the 1980s.
While a diversity of methods have been employed, it is unclear whether they provide
equivalent insights.
Here we review the most commonly used approaches for characterising and analysing morphological disparity, all of which have associated limitations that, if ignored, can lead to misinterpretation.
We provide best practice guidelines for disparity analyses, while noting that there can be no ``one-size-fits-all'' approach.
The available tools should always be used in the context of a specific biological question that will determine data and method selection at every stage of the analysis.

@@

Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids and molluscs), species-rich but phenotypically conservative (e.g. bacteria and nematodes), species-poor but phenotypically diverse (e.g. Afrotheria), or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not inextricably linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995)
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon. Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (Runnegar 1987, Gould \href{https://paperpile.com/c/sTGYvp/Uns3}{@@@}),
% Runnegar, Bruce "Rates and modes of evolution in the Mollusca." Rates of evolution. Allen and Unwin, London (1987): 39-60.
 this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

@@

(3) \emph{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
This makes disparity an informative complement to diversity measures based on species richness alone. Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling \href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(e.g., Fortey, Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and Neige 2007)}.
This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \href{https://paperpile.com/c/sTGYvp/dJHu}{(Losos 2011)}.

@@

(5) Disparity and evo-devo. Morphological disparity blalbalbal
% AG: Although intrinsic constraints are mentioned earlier, it doesn’t enter into any of these topics explicitly, but it is certainly one of the most interesting things one can use disparity for – to bridge development and evolution, looking at developmental biases/constraints. This also includes integration and modularity, which are also central to the topic of disparity and evolvability. Although you touch on these in places, I think it's really important to incorporate it into one of these topics, as there probably isn't room to give it one of its own (although i would argue it warrants equal footing with ecology, given that these are the processes that generate variation in the first place). There is a massive literature on using disparity to test hypotheses of developmental constraints, and also a huge one on disparity and integration/modularity. Let me know if you want me to send you refs.

Fundamental insights into evolutionary biology have been elicited from these five types of analysis.
One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \href{https://paperpile.com/c/sTGYvp/xxh5+yqPw}{(Hughes, Gerber, and Wills 2013; Foote 1997;} \href{https://paperpile.com/c/sTGYvp/EPJ2}{Erwin 2007)}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change \href{https://paperpile.com/c/sTGYvp/xxh5+Z6l6}{(Hughes, Gerber, and Wills 2013; Erwin 2011)}.
However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?

In attempting to answer this question, we review current methods and highlight their limitations, as part of a more general attempt to propose best practice guidelines for studies of disparity.
We first discuss the appropriate data required for characterising disparity, then review various challenging aspects of these approaches.
Throughout, it is important to remember these tools should always be used in the context of a specific scientific question, as this will drive data and methodological choices at every stage of the process.

@@

%@@@ Add disparity and evo-devo

@@

Disparity analyses are based on traits, but traits can be characterised in a number of ways:
% MH: We do not mention model-based descriptors at all, or the spaces defined by them (e.g. Raup's shell coiling models). I'm not convinced that we must, but it may come across as an notable omission to reviewers
% SG: I agree. There's for instance Saunders et al. (2004) disparity study of Paleozoic ammonoids using Raup's shell coiling parameters
(1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)};
(2) continuous measurements of features (e.g., lengths) \href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al. 2011)}; or
(3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, and Fourier coefficients \href{https://paperpile.com/c/sTGYvp/2Neu}{(Foote 1989;} \href{https://paperpile.com/c/sTGYvp/ZEDR}{Spriggs et al. 2018)}) (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits (Hetherington et al 2015; Hopkins 2017).

@@

The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
Discrete character data are commonly recycled from phylogenetic studies (e.g.Foote 1989\href{https://paperpile.com/c/sTGYvp/2Neu+0y4V}{; Deline et al. 2018}).
% SG: Foote (1989) is a study of trilobite disparity from Fourier analysis of cranidia. I don't think it has anything to do with discrete data or phylogenetic studies
% MH: Also the Foote papers that used discrete data weren't taken from phylogenetic studies but collected for the specific purpose of the project -- I think this might be true for the Deline paper as well. But there are many studies out there that used matrices initially intended for phylogenetics. I can look for some if needed.
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups \href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}.
This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}.
This effect of anatomical part on disparity patterns can be especially challenging when working on datasets where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al. 2018)}.

@@

Once suitable trait data have been collected, the design of the disparity analysis itself needs to be considered.
Study design encompasses several key aspects including (a) the difficulty of dealing with multidimensional data; (b) the variety of ordination (dimension reduction) techniques available and their limitations; (c) the metrics used to summarise the relative disparity of groups; (d) the methods used for hypothesis testing within the disparity analysis framework; and (e) ancestral state estimations in disparity-through-time analyses.
% @@@ Check the numbering here
We consider these aspects in turn below.
